---
title: "Origin of background RNA"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = "/data/share/htp/CZI_kidney/mouse_experiments/backgroundRNA")

library(tidyverse)
library(cowplot)
library(RVAideMemoire)

domesticus_colors = setNames(c("#271758", "#887BAF", "#C4BBDE", "lightgrey"),
                           c("endo", "empty", "cont", "other"))


profile_colors = setNames(c("red",  "purple", "orange", "lightgrey"),
                           c("endo", "empty", "cont", "other"))
```


## FIGURE 3: Characterization of ambient RNA in cells and empty droplets

### Figure 3A: UMI curve with definition of cells and empty droplets

UMI curve of total counts:

```{r UMI_curve, fig.height=4, fig.width=5}
raw_rep2 <- Seurat::Read10X_h5("Snakemake_backup/input/rep2/cellranger/raw_feature_bc_matrix.h5")
cells_rep2 <- read.table("Snakemake_backup/input/rep2/cell_BC.txt")$V1

assignment_rep2 <- data.frame(BC = colnames(raw_rep2), nUMI = Matrix::colSums(raw_rep2)) %>% 
    arrange(desc(nUMI)) %>% 
    mutate(index = row_number()) %>% 
    dplyr::filter(index <= 100000) %>% 
    mutate(isCell = ifelse(BC %in% cells_rep2, T, F))

lowest_cell <- assignment_rep2 %>% filter(isCell == T) %>% tail(1) %>% pull(index)

## UMI plot
p.UMI_plot <-  ggplot(assignment_rep2, aes(x = index, y = nUMI, color = isCell))+
    geom_point(size = 0.2, color = "black")+
    scale_y_log10(limits = c(10, 200000))+
    annotate(xmin = 0, xmax = lowest_cell, ymin = 0, ymax = Inf, geom = "rect", alpha = 0.1, fill = "red")+
    annotate(x = 9000, y = 180000, geom = "text", label = "cells")+
    annotate(xmin = 40000, xmax = 75000, ymin = 0, ymax = Inf, geom = "rect", alpha = 0.1, fill = "blue")+
    annotate(x = 58000, y = 180000, geom = "text", label = "empty droplets")+
    labs(x = "Droplet rank", y = "Total UMI counts")+
    theme_bw()+
    theme(legend.position = c(0.85,0.8))+ 
    guides(colour = guide_legend(override.aes = list(size=4)))
```

UMI curve of domesticus allele counts: 

```{r fig.width=5, fig.height=6}
csc_list <- list()

for(rep in c("rep1", "rep2", "rep3", "nuc2", "nuc3")){
#for(rep in c("rep2")){
  print(paste0("reading in replicate ", rep))
  csc_int <- readRDS(paste0("/data/share/htp/CZI_kidney/mouse_experiments/backgroundRNA/Snakemake_backup/genotype_estimation/",rep,"/csc_intermediate.RDS"))
  cell_assignment <- readRDS(paste0("/data/share/htp/CZI_kidney/mouse_experiments/backgroundRNA/deconvolution_new/raw_inputs/cell_assignment_",rep,".RDS"))
  csc_list[[rep]] <- csc_int %>% 
    filter(SNPisALT == "CAST") %>%
    dplyr::select(cell, position, DP, AD, ENSEMBL, SYMBOL) %>% 
    inner_join(cell_assignment, by = c("cell" = "BC"))
} 

## Strain proportions per barcode
perBC <- csc_list$rep2 %>% 
  group_by(cell_assignment, Strain, cell, nUMI) %>% 
  summarize(castaneus = sum(AD), 
            domesticus = sum(DP) - sum(AD),
            total = sum(DP)) %>% 
  mutate(category = case_when(Strain == "CAST" ~ "cont",
                              Strain %in% c("SvImJ","BL6") ~ "endo",
                              cell_assignment == "empty" ~ "empty",
                              T ~ "other"))  %>% 
  left_join(dplyr::select(assignment_rep2, BC, index), by = c("cell" = "BC"))

p.domesticus <- ggplot(perBC, aes(x = index, y = domesticus, color = category))+
    geom_point(size = 0.1, alpha = 0.1)+
    annotate(xmin = 0, xmax = lowest_cell, ymin = 0, ymax = Inf, geom = "rect", alpha = 0.1, fill = "red")+
    annotate(xmin = 40000, xmax = 75000, ymin = 0, ymax = Inf, geom = "rect", alpha = 0.1, fill = "blue")+
    #geom_point(aes(y = total), color = "black", size = 0.1)+
    scale_y_log10(limits = c(10,10000))+
    scale_color_manual(values = profile_colors)+
    scale_x_continuous(limits = c(0,100000))+
    theme_bw()+
    theme(legend.position = c(0.85,0.55), legend.title = element_blank())+
    labs(x = "Droplet rank", y = expression(paste("UMI counts (", italic("M.m.domesticus "), "alleles)")))+# "UMI counts (M.m.domesticus alleles)")+
    guides(colour = guide_legend(override.aes = list(size=3, alpha = 1)))

p.UMI_curve_combined <- plot_grid(p.UMI_plot+theme(axis.title.x = element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank(), plot.margin = unit(c(0, 0, 0, 0), "cm")), 
          p.domesticus+theme(plot.margin = unit(c(0, 0, 0, 0), "cm")), 
          nrow = 2, 
          align = "v", 
          rel_heights = c(1,1.2))
```

### Figure 3B: Correlation of pseudobulk profiles

```{r cor, fig.width=4, fig.height=3}
cor_list <- list()
for(rep in c("rep1", "rep2", "rep3", "nuc2", "nuc3")){
  prof <- readRDS(paste0("deconvolution_new/profile_comparisons/profiles_",rep,"_Gene_new.RDS"))$profiles_TMM
  for(profile in c("empty", "endo")){
    conf <- spearman.ci(prof[["cont"]], prof[[profile]], 1000, 0.95)
    cor_list[[rep]][[paste0("cont vs ", profile)]] <- data.frame(cor = conf$estimate, CI_low = conf$conf.int[1], CI_high = conf$conf.int[2])
  }
  cor_list[[rep]] <- bind_rows(cor_list[[rep]], .id = "profile")
}

cor_df <- bind_rows(cor_list, .id = "replicate") %>% 
  mutate(replicate = factor(replicate, levels = c("rep3","rep1","rep2","nuc3","nuc2")))

p.cor_flip <- cor_df %>% 
  mutate(profile = factor(profile, levels = c("cont vs empty", "cont vs endo")),
         replicate = factor(replicate, levels = rev(c("rep3","rep1","rep2","nuc3","nuc2")))) %>% 
  
  ggplot(aes(x = replicate, y = cor, fill = profile))+
  geom_col(color= "black", linewidth = 0.1, position = "dodge", width = 0.6)+
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0.2, position = position_dodge(width = 0.6))+
  coord_flip()+
  scale_fill_manual(values = c("darkgrey","lightgrey"))+
  theme_bw()+
  theme(axis.title.y = element_blank(), legend.title = element_blank(), legend.position = "right")+
  labs(y = expression("Spearman's "~rho), parse = T)

```

### Figure 3C: Deconvolutoin - cell type proportions

```{r deconvolution}
res.endo.dat <- readRDS("deconvolution_new/output/scdc/summaries/CZI_kidney_mouse_10XChromium_sc_CAST_singleref_endo_prop.rds")
res.cont.dat <- readRDS("deconvolution_new/output/scdc/summaries/CZI_kidney_mouse_10XChromium_sc_CAST_singleref_cont_prop.rds")
res.empty.dat <- readRDS("deconvolution_new/output/scdc/summaries/CZI_kidney_mouse_10XChromium_sc_CAST_singleref_empty_prop.rds")

res.qc <- readRDS("deconvolution_new/output/scdc/summaries/CZI_kidney_mouse_10XChromium_sc_CAST_singleref_deconvref.rds")
```

```{r combine_props}
all.endo<- res.endo.dat %>% 
  dplyr::select(Experiment, celltype, Pseudobulk, run, EstimatedProportion, TrueProportion) %>%
  dplyr::left_join(res.qc %>% dplyr::select(-cellnumber), 
                   by = c('Experiment' = 'Experiment', 'celltype' = 'celltype')) %>% 
  tidyr::pivot_longer(cols = -c(celltype, Experiment, Pseudobulk, run), names_to = "Type", values_to = 'Proportion') %>% 
  dplyr::mutate(run=ifelse(is.na(run), "1", run)) %>% 
  dplyr::mutate(Type=case_when(Type == 'EstimatedProportion' ~ 'Estimated',
                               Type == 'TrueProportion' ~ 'True',
                               Type == 'refproportion' ~ 'Reference')) %>% 
  dplyr::filter(Pseudobulk == 'all') 



all.cont<- res.cont.dat %>% 
  dplyr::select(Experiment, celltype, Pseudobulk, run, EstimatedProportion) %>%
  tidyr::pivot_longer(cols = -c(celltype, Experiment, Pseudobulk, run), names_to = "Type", values_to = 'Proportion') %>% 
  dplyr::mutate(run=ifelse(is.na(run), "1", run)) %>% 
  dplyr::mutate(Type=case_when(Type == 'EstimatedProportion' ~ 'Contamination')) %>% 
  dplyr::filter(Pseudobulk == 'all') 



# estimated proportion: bar plot of all cells
all.empty <- res.empty.dat %>% 
  dplyr::select(Experiment, celltype, Pseudobulk, rprop, run, EstimatedProportion) %>%
  dplyr::filter(Pseudobulk == 'subsample' & rprop == "1") %>% 
  dplyr::select(-Pseudobulk, -rprop, -run) %>% 
  tidyr::pivot_longer(cols = -c(celltype, Experiment), names_to = "Type", values_to = 'Proportion') %>% 
  dplyr::mutate(Type=case_when(Type == 'EstimatedProportion' ~ 'Empty'))

all.props<-bind_rows(all.endo,all.cont, all.empty) %>%
  dplyr::mutate(Type = factor(Type, levels = c("Reference","True", "Estimated", "Contamination", "Empty")), 
                replicate = case_when(Experiment == 1 ~ "rep1",
                                      Experiment == 2 ~ "rep2",
                                      Experiment == 3 ~ "rep3", 
                                      Experiment == 4 ~ "nuc2",
                                      Experiment == 5 ~ "nuc3"),
                replicate = factor(replicate, levels = c("rep3","rep1","rep2","nuc3","nuc2")))
```

Plot proportions: 

```{r plot_prop}
celltype_colors <- setNames(
  c(RColorBrewer::brewer.pal(11, "Paired"),"darkgrey","grey"),
  c("PT","CD_IC" ,   "CD_PC", "CD_Trans" ,     "CNT" ,     "DCT",     "Endo",      "Fib","aLOH",   "dLOH",    "MC" ,    "Podo","Immune"))

# Two example replicates: 
p.deconv <- all.props %>% 
  filter(Type %in% c("Estimated", "Empty", "Contamination"),
         replicate %in% c("rep2", "nuc2")) %>% 
  mutate(category = case_when(Type == "Empty" ~ "empty",
                              Type == "Contamination" ~ "cont",
                              Type == "Estimated" ~ "endo")) %>% 

  ggplot(aes(x = category, y = Proportion, fill = forcats::fct_rev(celltype))) + 
    geom_col(color = 'black', lwd=0.1) + 
    scale_fill_manual(values = celltype_colors) +
    guides(fill = guide_legend(nrow = 2)) +
    scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) +
    coord_flip() +
    facet_grid(~replicate, scales='free') +
    theme_bw() +
    theme(legend.position = 'bottom',
          axis.title.y = element_blank(),
          legend.title = element_blank(),
          legend.text = element_text(size = 8),
          legend.key.size = unit(0.5, 'cm'),
          axis.text = element_text(size = 10),
          axis.title = element_text(size = 12, face='bold'),
          strip.text = element_text(size = 12, face = 'bold'),
          strip.background = element_blank()) 
```

### Figure 1D: Intron fractions

```{r intron_profile}
SNP_matrix_list <- readRDS("Analysis/files/SNP_matrix_list_check.RDS")

calculate_intron_fractions <- function(matrices, vcf){
  data.frame(fraction = c(sum(matrices$empty[rownames(matrices$empty) %in% vcf$position[vcf$intron_variant == T], ]) / sum(matrices$empty),
             sum(matrices$cont[rownames(matrices$cont) %in% vcf$position[vcf$intron_variant == T], ]) / sum(matrices$cont),
             sum(matrices$endo[rownames(matrices$endo) %in% vcf$position[vcf$intron_variant == T], ]) / sum(matrices$endo)),
             profile = c("empty", "cont", "endo"))
}

intron_frac <-  lapply(SNP_matrix_list, function(x){
   calculate_intron_fractions(matrices = x$matrices, vcf = x$vcf) 
  }) %>% bind_rows(.id = "replicate")


p.intron_frac <- intron_frac %>% 
  mutate(profile = factor(profile, levels = c("empty", "cont", "endo")),
         replicate = factor(replicate, levels = rev(c("rep3","rep1","rep2","nuc3","nuc2")))) %>% 
  ggplot(aes(x = replicate, y = fraction, color = profile))+
  coord_flip()+
  geom_jitter(size = 3, width = 0.2)+
  scale_color_manual(values = profile_colors, limits = force)+
  theme_bw()+
  labs(y = "fraction of intronic variants", x = "")+
  theme(legend.title = element_blank(), axis.title.y = element_blank())
```

### Assemble Figure 3

```{r figure, fig.width=12, fig.height=5.5}
p.figure4 <- plot_grid(
  p.UMI_curve_combined,
  plot_grid(
    plot_grid(p.cor_flip+theme(legend.position = "none"),
            p.intron_frac+theme(legend.position = "none", legend.title= element_blank()),
            plot_grid(NULL,
                      get_legend(p.cor_flip),#+theme(legend.justification=c(0,0)), 
                      get_legend(p.intron_frac),#+theme(legend.justification=c(0,0)),
                      NULL,
                      ncol = 1, align = "v", rel_heights = c(0.4,1,1,0.2)), 
            nrow = 1, rel_widths = c(1,1,0.63), align = "hv", labels = c("B", "D")),
    p.deconv+theme(legend.position = "right")+guides(fill = guide_legend(ncol = 2)),
    nrow = 2, labels = c("","C")), 
  nrow = 1, rel_widths = c(1,1.7), scale = c(0.9,1), labels = c("A","")
)

p.figure4

#ggsave("figures/Figure3_cont_origin.png", p.figure4, height = 5.5, width = 12)
```

## SUPPL. FIGURE S4: UMI curves for all replicates

```{r UMI_curve_replicates}
plot_UMI_curves <- function(rep, empty_min, empty_max, no_y_lab = F){
  raw_mat <- Seurat::Read10X_h5(paste0("Snakemake_backup/input/",rep,"/cellranger/raw_feature_bc_matrix.h5"))
  cell_BC <- read.table(paste0("Snakemake_backup/input/",rep,"/cell_BC.txt"))$V1
  
  assignment <- data.frame(BC = colnames(raw_mat), nUMI = Matrix::colSums(raw_mat)) %>% 
      arrange(desc(nUMI)) %>% 
      mutate(index = row_number()) %>% 
      dplyr::filter(index <= 100000) %>% 
      mutate(isCell = ifelse(BC %in% cell_BC, T, F))
  
  lowest_cell <- assignment %>% filter(isCell == T) %>% tail(1) %>% pull(index)
  
  ## UMI plot
  p.UMI_plot <- assignment %>% 
      ggplot(aes(x = index, y = nUMI, color = isCell))+
        geom_point(size = 0.2, color = "black")+
        scale_y_log10(limits = c(10, 200000))+
        annotate(xmin = 0, xmax = lowest_cell, ymin = 0, ymax = Inf, geom = "rect", alpha = 0.1, fill = "red")+
        annotate(x = 9000, y = 180000, geom = "text", label = "cells")+
        annotate(xmin = empty_min, xmax = empty_max, ymin = 0, ymax = Inf, geom = "rect", alpha = 0.1, fill = "blue")+
        annotate(x = 58000, y = 180000, geom = "text", label = "empty droplets")+
        labs(x = "Droplet rank", y = "Total UMI counts")+
        theme_bw()+
        theme(legend.position = c(0.85,0.8))+ 
        guides(colour = guide_legend(override.aes = list(size=4)))+
        ggtitle(rep)
  
  print(paste0("summarizing allele counts in replicate ", rep))
  perBC <- csc_list[[rep]] %>% 
    group_by(cell_assignment, Strain, cell, nUMI) %>% 
    dplyr::summarize(castaneus = sum(AD), 
              domesticus = sum(DP) - sum(AD),
              total = sum(DP)) %>% 
    mutate(category = case_when(Strain == "CAST" ~ "cont",
                                Strain %in% c("SvImJ","BL6") ~ "endo",
                                cell_assignment == "empty" ~ "empty",
                                T ~ "other"))  %>% 
    left_join(dplyr::select(assignment, BC, index), by = c("cell" = "BC"))
  
  p.domesticus <- ggplot(perBC, aes(x = index, y = domesticus, color = category))+
      geom_point(size = 0.1, alpha = 0.1)+
      annotate(xmin = 0, xmax = lowest_cell, ymin = 0, ymax = Inf, geom = "rect", alpha = 0.1, fill = "red")+
      annotate(xmin = empty_min, xmax = empty_max, ymin = 0, ymax = Inf, geom = "rect", alpha = 0.1, fill = "blue")+
      #geom_point(aes(y = total), color = "black", size = 0.1)+
      scale_y_log10(limits = c(10,10000))+
      scale_color_manual(values = profile_colors)+
      scale_x_continuous(limits = c(0,100000))+
      theme_bw()+
      theme(legend.position = c(0.85,0.55), legend.title = element_blank())+
      labs(x = "Droplet rank", y = expression(paste("UMI counts (", italic("M.m.domesticus "), "alleles)")))+# "UMI counts (M.m.domesticus alleles)")+
      guides(colour = guide_legend(override.aes = list(size=3, alpha = 1)))
  
  if(isTRUE(no_y_lab)){
    p.UMI_plot <- p.UMI_plot+labs(y = "")
    p.domesticus <- p.domesticus+labs(y = "")
  }
  
  p.UMI_curve_combined <- plot_grid(p.UMI_plot+theme(axis.title.x = element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank(), plot.margin = unit(c(0, 0, 0, 0), "cm")), 
            p.domesticus+theme(plot.margin = unit(c(0, 0, 0, 0), "cm")), 
            nrow = 2, 
            align = "v", 
            rel_heights = c(1,1.2))
  
  return(p.UMI_curve_combined)
}

UMI_curve_list <- list()
UMI_curve_list$rep1 = plot_UMI_curves("rep1", empty_min = 40000, empty_max = 75000)
UMI_curve_list$rep2 = plot_UMI_curves("rep2", empty_min = 40000, empty_max = 75000, no_y_lab = T)
UMI_curve_list$rep3 = plot_UMI_curves("rep3", empty_min = 20000, empty_max = 60000, no_y_lab = T)
UMI_curve_list$nuc2 = plot_UMI_curves("nuc2", empty_min = 30000, empty_max = 70000, no_y_lab = T)
UMI_curve_list$nuc3 = plot_UMI_curves("nuc3", empty_min = 20000, empty_max = 60000, no_y_lab = T)

p.UMI_curve_replicates <- plot_grid(plotlist = UMI_curve_list, nrow = 1 )

p.UMI_curve_replicates

#ggsave("figures/Suppl_figure_UMI_curves.png", p.UMI_curve_replicates, height = 5.5, width = 20)
```

## SUPPL. FIGURE S5: Deconvolution - cell type proportions for all replicates

```{r deconvolution_all_rep, fig.width=10, fig.height=3.6}
p.prop_allRep <- all.props %>% 
  filter(Type %in% c("Reference","Estimated", "Empty", "Contamination")) %>% 
  mutate(category = case_when(Type == "Empty" ~ "empty",
                              Type == "Contamination" ~ "cont",
                              Type == "Estimated" ~ "endo",
                              Type == "Reference" ~ "reference")) %>% 

  ggplot(aes(x = category, y = Proportion, fill = forcats::fct_rev(celltype))) + 
    geom_col(color = 'black', lwd=0.1) + 
    scale_fill_manual(values = celltype_colors) +
    guides(fill = guide_legend(nrow = 2)) +
    scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) +
    coord_flip() +
    facet_grid(~replicate, scales='free') +
    theme_bw() +
    theme(legend.position = 'bottom',
          axis.title.y = element_blank(),
          legend.title = element_blank(),
          legend.text = element_text(size = 8),
          legend.key.size = unit(0.5, 'cm'),
          axis.text.y = element_text(size = 10),
          axis.text.x = element_text(size = 8),
          axis.title = element_text(size = 12, face='bold'),
          strip.text = element_text(size = 12, face = 'bold'),
          strip.background = element_blank()) 

p.prop_allRep

#ggsave("figures/Suppl_figure_Deconvolution_allRep.pdf", p.prop_allRep, width = 10, height = 3.6)
```
