---
title: "Benchmark: downstream analysis"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = "/data/share/htp/CZI_kidney/mouse_experiments/backgroundRNA")

library(tidyverse)
library(tidyseurat)
library(cowplot)
library(ggh4x)

method_colors2 = setNames(c("lightgrey", "lightgrey","#FF6F59", "#254441", "#43AA8B", "#9EEED2"),
                         c("Genotype estimate", "raw", "CellBender", "SoupX", "DecontX_default", "DecontX_background"))

replicate_colors <- setNames(c("#F6D6B0","#917AA6","#D4E7A5", "lightgrey"), 
                             c("rep1", "rep2", "rep3", "nuc2")) 

denisenko_colors <- setNames(
 c("#88CCEE", "#6699CC",  "#44AA99","#117733", "#CC6677",  "#661100","#DDCC77", "#882255", "#AA4499", "#332288", "#999933", "#888888","lightgrey"),
  c("PT",      "CD_IC" ,   "CD_PC", "CD_Trans" , "CNT" ,    "DCT",     "Endo",    "Fib",     "aLOH",   "dLOH",     "MC" ,    "Podo",   "Immune"))
```


## FIGURE 6: EFFECT OF BACKGROUND REMOVAL ON DOWNSTREAM ANALYSIS

### Figure 6A: UMAP with cell type annotation before and after correction

```{r prepare_seurat}
prepare_seurat <- function(rep){
  cnt_mat_list <- list(uncorrected = readRDS(paste0("Snakemake_backup/benchmark/corrected/raw/",rep,"/raw_cormat.RDS")),
                       CellBender = readRDS(paste0("Snakemake_backup/benchmark/corrected/CellBender/",rep,"/fpr0.01_total25000_cormat.RDS")),
                       SoupX = readRDS(paste0("Snakemake_backup/benchmark/corrected/SoupX/",rep,"/res1_contAuto_cormat.RDS")),
                       DecontX = readRDS(paste0("Snakemake_backup/benchmark/corrected/DecontX/",rep,"/resDefault_emptyFalse_cormat.RDS")))
  
  class_list <- list(uncorrected = readRDS(paste0("Snakemake_backup/benchmark/evaluation/raw/",rep,"/classification_evaluation.RDS"))$raw$celltype_labels,
                     CellBender = readRDS(paste0("Snakemake_backup/benchmark/evaluation/CellBender/",rep,"/classification_evaluation.RDS"))$fpr0.01_total25000$celltype_labels,
                     SoupX = readRDS(paste0("Snakemake_backup/benchmark/evaluation/SoupX/",rep,"/classification_evaluation.RDS"))$res1_contAuto$celltype_labels,
                     DecontX = readRDS(paste0("Snakemake_backup/benchmark/evaluation/DecontX/",rep,"/classification_evaluation.RDS"))$resDefault_emptyFalse$celltype_labels)
  
  
  seu_list <- lapply(setNames(names(cnt_mat_list), names(cnt_mat_list)), function(method){
    mat_filt <- cnt_mat_list[[method]][,class_list[[method]]$cell]
    seu <- CreateSeuratObject(mat_filt, meta.data = class_list[[method]] %>% column_to_rownames("cell"))
    #seu <- filter(seu, !is.na(celltype))
    seu <- NormalizeData(seu)
    seu <- FindVariableFeatures(seu, nfeatures = 5000)
    seu <- ScaleData(seu)
    seu <- RunPCA(seu)
    seu <- RunUMAP(seu, dims = 1:30)
    seu <- seu %>% mutate(celltype = case_when(celltype %in% c("CD_IC","CD_IC_A","CD_IC_B") ~ "CD_IC",
                                #celltype %in% c("aLOH","dLOH") ~ "LOH",
                                celltype %in% c("T","NK","B","MPH") ~ "Immune",
                                T ~ celltype))
    return(seu)
  })
  
  seu_list <- lapply(setNames(names(seu_list),names(seu_list)), function(x){
    seu_list[[x]]$label_change = ifelse(seu_list[[x]]$celltype != seu_list$uncorrected$celltype,x,"same")
    seu_list[[x]]$original_label = seu_list$uncorrected$celltype
    return(seu_list[[x]])
  })
  
  
  seu_list$uncorrected$label_change <- case_when(
    seu_list$uncorrected$celltype != seu_list$CellBender$celltype & seu_list$uncorrected$celltype != seu_list$DecontX$celltype ~ "multiple",
    seu_list$uncorrected$celltype != seu_list$CellBender$celltype & seu_list$uncorrected$celltype != seu_list$SoupX$celltype ~ "multiple",
    seu_list$uncorrected$celltype != seu_list$SoupX$celltype & seu_list$uncorrected$celltype != seu_list$DecontX$celltype ~ "multiple",
    seu_list$uncorrected$celltype != seu_list$CellBender$celltype & seu_list$uncorrected$celltype != seu_list$SoupX$celltype & seu_list$uncorrected$celltype != seu_list$DecontX$celltype ~ "multiple",
    seu_list$uncorrected$celltype != seu_list$CellBender$celltype ~ "CellBender",
    seu_list$uncorrected$celltype != seu_list$SoupX$celltype ~ "SoupX",
    seu_list$uncorrected$celltype != seu_list$DecontX$celltype ~ "DecontX",
    T ~ "same"
  )
  
  return(seu_list)
}

seu_list_rep1 <- prepare_seurat("rep1")
seu_list_rep2 <- prepare_seurat("rep2")
seu_list_rep3 <- prepare_seurat("rep3")
seu_list_nuc2 <- prepare_seurat("nuc2")
seu_list_nuc3 <- prepare_seurat("nuc3")

```

Plot UMAPs for replicate 2: 

```{r plot_UMAPs_rep2}
plot_UMAP <- function(seurat, color_by = "celltype", colors, plot_legend = F, name, highlight_label_change = T, add_caption = T){
  p <- ggplot(filter(seurat, !is.na(celltype)), aes_string(x = "UMAP_1", y = "UMAP_2", color = color_by))+
    geom_point(size = 0.1)+
    scale_color_manual(values = colors)+
    theme_bw()+
    facet_grid(~as.character(name))+
    theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(),
          strip.text = element_text(size = 9))
  
  if(plot_legend == F){
    p <- p+theme(legend.position = "none")
  }
  
  if(highlight_label_change == T){
    if(sum(seurat$label_change != "same", na.rm = T) > 0){
      p <- p+geom_point(data = filter(as.tibble(seurat), label_change != "same"), mapping =  aes_string(x = "UMAP_1", y = "UMAP_2",fill = color_by, shape = "label_change"), size = 1.3, color = "black")+
        scale_fill_manual(values = colors)+
        scale_shape_manual(values = setNames(c(21,22,23,24),c("multiple", "CellBender","SoupX","DecontX")))
    }
  }
  
  if(add_caption == T){
    n_reassigned <- sum(seurat$label_change != "same", na.rm = T)
    frac_reassigned <- sum(seurat$label_change != "same", na.rm = T) / length(seurat$label_change)
    label_reassigned <- paste0("reassigned: ", n_reassigned, " cells (",round(frac_reassigned*100,1),"%)")
    if(name == "uncorrected"){label_reassigned = ""}
    p <- p + labs(caption = label_reassigned)
  }
  
  return(p)
}

p.UMAPs_ct <- plot_grid(plotlist = lapply(setNames(names(seu_list_rep2), names(seu_list_rep2)),
                            function(x){plot_UMAP(seu_list_rep2[[x]], colors = denisenko_colors,name = x)}),
          nrow = 1)
```

### Figure 6B: UMAP with Slc34a1 expression before and after correction

```{r UMAP_marker_rep2}
plot_UMAP_marker <- function(seu, gene, name){
  expr <- seu@assays$RNA@counts[gene,]
  seu$log_expr <- log2(expr+1)
  seu$log_expr <- ifelse(seu$log_expr > 6, 6, seu$log_expr)
  
  p.UMAP <- ggplot(filter(seu, !is.na(log_expr)), aes(x =UMAP_1, y = UMAP_2, color = log_expr))+
      geom_point(size = 0.1)+
      scale_color_continuous(low = "lightgrey", high = "red",limits = c(0,6))+
      #scale_y_continuous(limits = c(-20,15))+
      #annotate("text", x = 3, y = 8, label = "PT cells")+
      theme_bw()+
      facet_grid(~as.character(name))+
      #ggtitle("Slc34a1 expression (rep 2)")+
      labs(x = "UMAP 1", y = "UMAP 2", color = "log2(count+1)")+
      theme(legend.position = "none", 
            axis.text = element_blank(), axis.ticks = element_blank(), axis.title = element_blank(),
            strip.text = element_text(size = 10))
  
    if(sum(seu$label_change != "same", na.rm = T) > 0){
      p.UMAP <- p.UMAP+geom_point(data = filter(as.tibble(seu), label_change != "same" & original_label == "PT"), mapping =  aes(fill = log_expr, shape = label_change), size = 1.3, color = "black")+
        scale_shape_manual(values = setNames(c(21,22,23,24),c("multiple", "CellBender","SoupX","DecontX")))+
        scale_fill_continuous(low = "lightgrey", high = "red",limits = c(0,6))
    }
  
  return(p.UMAP)
}

p.UMAPs_marker <- plot_grid(plotlist = lapply(setNames(names(seu_list_rep2), names(seu_list_rep2)),
                            function(x){plot_UMAP_marker(seu_list_rep2[[x]], gene = "Slc34a1",name = x)+
                                theme(strip.background = element_blank(),strip.text.x = element_blank())
                                }),
          nrow = 1)
```

### Figure 6C: Expression fraction

```{r expression_frac}
topGenes <- readRDS("Extra_analyses/top10_PT_markers.RDS")

l_expr_perGene <- list()
for(method in c("raw", "CellBender", "DecontX", "SoupX")){
  for(rep in c("rep1","rep2","rep3","nuc2")){ 
    expression_fraction_PT_markers <- readRDS(paste0("Snakemake_backup/benchmark/evaluation/",method,"/",rep,"/expression_fraction_PT_markers.RDS"))
    l_expr_perGene[[method]][[rep]] <- expression_fraction_PT_markers 
  }
  l_expr_perGene[[method]] <- bind_rows(l_expr_perGene[[method]], .id = "replicate")
} 

df_expr_perGene <- bind_rows(l_expr_perGene, .id = "method")

p.expr_frac_topGenes <- df_expr_perGene %>% 
  mutate(default = (method == "CellBender" & param == "fpr0.01_total25000" |
                    method == "DecontX" & param == "resDefault_emptyFalse" |
                    method == "DecontX" & param == "resDefault_emptyTrue" |
                    method == "SoupX" & param == "res1_contAuto" |
                    method == "raw"),
         replicate = factor(replicate, levels = c("rep3", "rep1", "rep2", "nuc3", "nuc2")),
         method2 = case_when(method == "DecontX" & param == "resDefault_emptyFalse" ~ "DecontX_default",
                             method == "DecontX" & param == "resDefault_emptyTrue" ~ "DecontX_background",
                             T ~ method))  %>% 
  #left_join(dplyr::select(df_eval_metrics, c(method,replicate, param,default))) %>% 
  #distinct() %>% 
  filter(gene %in% topGenes, default == T) %>% 
  mutate(gene = factor(gene, levels = rev(topGenes)),
         replicate = factor(replicate, levels = rev(c("rep3", "rep1", "rep2", "nuc2"))),
         method2 = factor(method2, levels = rev(c("raw", "CellBender","SoupX", "DecontX_default", "DecontX_background")))) %>%

  ggplot(aes(x = replicate, y = expr_within_others, fill = method2))+
    geom_boxplot(alpha = 0.8, outlier.size = 0.1)+
    scale_fill_manual(values = method_colors2, limits = force, breaks = c("raw", "CellBender","SoupX", "DecontX_default", "DecontX_background"))+
    coord_flip()+
    theme_bw()+
    theme(axis.title.y = element_blank(), legend.position = "none")+
    labs(y = "Expression fraction in non-PT cells")
```

### Figure 6D: Log2 fold changes

```{r log2fc}
l_lfc_perGene_seurat <- list()
for(method in c("raw", "CellBender", "DecontX", "SoupX")){
  for(rep in  c("rep1", "rep2","rep3", "nuc2")){ 
    differential_expression <- readRDS(paste0("Snakemake_backup/benchmark/evaluation/",method,"/",rep,"/differential_expression_seurat.RDS"))
    l_lfc_perGene_seurat[[method]][[rep]] <- differential_expression %>% data.frame %>% 
       rownames_to_column("gene") %>% 
       mutate(gene = gsub("\\..*", "", gene)) %>% 
       dplyr::filter(gene %in% topGenes)
  }
  l_lfc_perGene_seurat[[method]] <- bind_rows(l_lfc_perGene_seurat[[method]], .id = "replicate")
} 

df_lfc_perGene_seurat <- bind_rows(l_lfc_perGene_seurat, .id = "method")

p.seurat_lfc_topGenes <- df_lfc_perGene_seurat %>% 
  mutate(default = (method == "CellBender" & param == "fpr0.01_total25000" |
                    method == "DecontX" & param == "resDefault_emptyFalse" |
                    method == "DecontX" & param == "resDefault_emptyTrue" |
                    method == "SoupX" & param == "res1_contAuto" |
                    method == "raw"),
         replicate = factor(replicate, levels = c("rep3", "rep1", "rep2", "nuc3", "nuc2")),
         method2 = case_when(method == "DecontX" & param == "resDefault_emptyFalse" ~ "DecontX_default",
                             method == "DecontX" & param == "resDefault_emptyTrue" ~ "DecontX_background",
                             T ~ method))  %>% 
  filter(gene %in% topGenes, default == T) %>% 
  mutate(gene = factor(gene, levels = rev(topGenes)),
         replicate = factor(replicate, levels = rev(c("rep3", "rep1", "rep2", "nuc2"))),
         method2 = factor(method2, levels = rev(c("raw", "CellBender","SoupX", "DecontX_default", "DecontX_background")))) %>%

  ggplot(aes(x = replicate, y = avg_log2FC, fill = method2))+
    geom_boxplot(alpha = 0.8, outlier.size = 0.1)+
    scale_fill_manual(values = method_colors2, limits = force, breaks = c("raw", "CellBender","SoupX", "DecontX_default", "DecontX_background"))+
    coord_flip()+
    theme_bw()+
    theme(axis.title.y = element_blank(), legend.position = "none")+
    labs(y = "log2FC")
```

### Figure 6E: Clustering and classification

```{r cluster_and_classification}
df_eval_metrics <- readRDS("Extra_analyses/benchmark_metrics.RDS")

inp <- df_eval_metrics %>% 
  filter(default == T,
         metric %in% c("tau", "rmsle","prediction_score", "purity", "kNN_overlap", "avg_silhouette","log_ratio_expression", "lfc")) %>% 
  mutate(evaluation_category = factor(evaluation_category, levels = c("estimation_accuracy", "cluster_evaluation", "marker_evaluation")),
         method = factor(method, levels = c("CellBender","SoupX","DecontX","raw")),
         method2 = factor(method2, levels = c("CellBender","SoupX","DecontX_default","DecontX_background","raw")),
         metric_long = case_when(metric == "avg_silhouette" ~ "Average silhouette",
                                 metric == "prediction_score" ~ "Prediction score (delta)",
                                 metric == "purity" ~ "Purity",
                                 metric == "kNN_overlap" ~ "k-NN overlap",
                                 T ~ metric))

## scaled by range
p.cluster_evaluation_range <- inp %>% 
  mutate(scaled_range = case_when(metric %in% c("avg_silhouette","prediction_score") ~ delta/2,
                                  metric %in% c("purity") ~ delta,
                                  metric %in% c("kNN_overlap") ~ delta/50)) %>% 
  filter(method != "raw", evaluation_category == "cluster_evaluation") %>% 
  mutate(metric_long = factor(metric_long, levels = c("Prediction score (delta)", "Average silhouette","Purity","k-NN overlap")),
         replicate = factor(replicate, levels = rev(c("rep3", "rep1", "rep2", "nuc2")))) %>% 
  ggplot(aes(x = replicate, y = scaled_range, color = method2))+
      geom_jitter(size = 2, width = 0.1)+
      geom_hline(data = filter(inp, evaluation_category != "estimation_accuracy"), yintercept = 0, color = "black")+
      coord_flip()+
      #scale_y_continuous(limits = c(-0.5,0.5))+
      scale_color_manual(values = method_colors2, limits = force)+
      ggh4x::facet_nested(~metric_long)+
      #ggh4x::facetted_pos_scales(y = list(metric == "rmsle" ~ scale_y_reverse()))+
      theme_bw()+
      labs(y = "scaled difference to score(uncorrected)")+
          theme(axis.title.y = element_blank(), axis.text.x = element_text(size = 6), strip.background = element_rect(fill = "white"))
```

### Assemble Figure 6

```{r assemble_fig6, fig.width=9, fig.height=9}
legend_celltypes_wide <- get_legend(ggplot(filter(seu_list_rep2$CellBender, !is.na(celltype)), aes(x=UMAP_1, y = UMAP_2, color = celltype))+
  geom_point()+theme_bw()+guides(colour = guide_legend(override.aes = list(size=2), nrow = 2))+theme(legend.position = "top")+
    scale_color_manual(values = denisenko_colors, limits = force))

legend_reassignment <- get_legend(ggplot(data = filter(seu_list_rep2$uncorrected, label_change != "same"), mapping =  aes_string(x = "UMAP_1", y = "UMAP_2",shape = "label_change"), size = 3, color = "black")+ geom_point()+
        scale_shape_manual(values = setNames(c(21,22,23,24),c("multiple", "CellBender","SoupX","DecontX")))+
        guides(shape=guide_legend(title="reassignment in"))+
        theme_bw())
                                
legend_Slc_expr_wide <- get_legend(plot_UMAP_marker(filter(seu_list_rep2$uncorrected, label_change != "same"), gene = "Slc34a1",name = "uncorrected")+
                                theme(legend.position = "bottom")+labs(color = "Slc34a1 expr. \n [log2(counts+1)]")+
                                  guides(fill = "none", shape= "none"))

p.UMAPs <- plot_grid(
  plot_grid(
    legend_celltypes_wide,
    p.UMAPs_ct, 
    p.UMAPs_marker,
    legend_Slc_expr_wide,
    nrow = 4, labels = c("","A","B",""), scale = c(1,0.95,0.95,1), rel_heights = c(0.2,1.1,0.9,0.2)),
  legend_reassignment,
  nrow = 1, rel_widths = c(1,0.15)
)

p.downstream_benchmark_new <- plot_grid(
  p.UMAPs,
  plot_grid(
    p.expr_frac_topGenes,
    p.seurat_lfc_topGenes+theme(legend.position = "right", legend.title = element_blank()),
    nrow = 1, labels = c("C","D"), scale = c(0.95,0.95), align = "hv", rel_widths = c(0.65,1)#, axis = "blrt"
  ),
  plot_grid(p.cluster_evaluation_range+theme(legend.position = "none"), labels = "E"),
  nrow = 3, rel_heights = c(1,0.6,0.6), scale = c(0.95,0.95,0.95), align = "hv", axis = "l")

p.downstream_benchmark_new

ggsave("figures/figures_final/Figure6.pdf", p.downstream_benchmark_new, width = 9, height = 9)
```

## SUPPL. FIGURE S11: cell type UMAP all replicates

```{r UMAPs_ct, fig.width=9, fig.height=12}
plot_UMAP_ct <- function(seu_list){
   plot_grid(plotlist = lapply(setNames(names(seu_list), names(seu_list)),
                            function(x){plot_UMAP(seu_list[[x]], colors = denisenko_colors,name = x)}),
          nrow = 1)
}

legend_celltypes_wide2 <- get_legend(ggplot(filter(seu_list_rep2$CellBender, !is.na(celltype)), aes(x=UMAP_1, y = UMAP_2, color = celltype))+
  geom_point()+theme_bw()+guides(colour = guide_legend(override.aes = list(size=2), nrow = 2))+
    scale_color_manual(values = denisenko_colors, limits = force))
legend_reassignment2 <- get_legend(ggplot(data = filter(seu_list_rep2$uncorrected, label_change != "same"), mapping =  aes_string(x = "UMAP_1", y = "UMAP_2",shape = "label_change"), size = 3, color = "black")+ geom_point()+
        scale_shape_manual(values = setNames(c(21,22,23,24),c("multiple", "CellBender","SoupX","DecontX")))+
        guides(shape=guide_legend(title="reassignment in", nrow = 2))+ 
        theme_bw())


p.UMAP_ct_rep <- plot_grid(
  NULL,
  plot_UMAP_ct(seu_list_rep1),
  plot_UMAP_ct(seu_list_rep2),
  plot_UMAP_ct(seu_list_rep3),
  plot_UMAP_ct(seu_list_nuc2),
  plot_UMAP_ct(seu_list_nuc3), 
  plot_grid(legend_celltypes_wide2, legend_reassignment2),
  ncol = 1, labels = c("","rep1", "rep2", "rep3", "nuc2", "nuc3"), label_y = 1.11, rel_heights = c(0.1,1,1,1,1,1,0.35)
)

p.UMAP_ct_rep

ggsave("figures/figures_final/Suppl_figure_S11_UMAPs_corrected_CT.pdf",  p.UMAP_ct_rep, width = 9, height = 12)
```

## SUPPL. FIGURE S12: Slc34a1 UMAPs all replicates

```{r UMAPs_slc, fig.width=9, fig.height=11}
plot_UMAP_corrected_slc <- function(seu_list){
   plot_grid(plotlist = lapply(setNames(names(seu_list), names(seu_list)),
                            function(x){plot_UMAP_marker(seu_list[[x]], gene = "Slc34a1",name = x)
                                }),
          nrow = 1)
}

p.UMAP_slc_rep <- plot_grid(
  NULL,
  plot_UMAP_corrected_slc(seu_list_rep1),
  plot_UMAP_corrected_slc(seu_list_rep2),
  plot_UMAP_corrected_slc(seu_list_rep3),
  plot_UMAP_corrected_slc(seu_list_nuc2),
  plot_UMAP_corrected_slc(seu_list_nuc3), 
  plot_grid(legend_Slc_expr_wide, legend_reassignment2),
  ncol = 1, labels = c("","rep1", "rep2", "rep3", "nuc2", "nuc3"), label_y = 1.07, rel_heights = c(0.1,1,1,1,1,1,0.35), scale = 0.95
)

p.UMAP_slc_rep

ggsave("figures/figures_final/Suppl_figure_S12_UMAPs_corrected_Slc34a1.pdf",  p.UMAP_slc_rep, width = 9, height = 11)
```

## SUPPL. FIGURE S13: Effect on marker gene detection

### S13A: Heatmap

```{r marker_heatmap}
cell_metadata <- readRDS("Extra_analyses/cell_metadata.RDS")

cell_sample <- cell_metadata %>% 
  filter(Strain == "CAST", !is.na(celltype), replicate != "nuc3") %>% 
  mutate(isPT = ifelse(celltype == "PT", "PT", "not PT")) %>% 
  group_by(replicate, isPT) %>% 
  slice_sample(n = 100)

marker_plot_inp <- lapply(setNames(c("rep1", "rep2","rep3", "nuc2"), c("rep1", "rep2","rep3", "nuc2")), function(rep){ 
  cormat_list <- list(raw = readRDS(paste0("Snakemake_backup/benchmark/corrected/raw/",rep,"/raw_cormat.RDS")),
                      CellBender = readRDS(paste0("Snakemake_backup/benchmark/corrected/CellBender/",rep,"/fpr0.01_total25000_cormat.RDS")),
                      DecontX_default = readRDS(paste0("Snakemake_backup/benchmark/corrected/DecontX/",rep,"/resDefault_emptyFalse_cormat.RDS")),
                      DecontX_background = readRDS(paste0("Snakemake_backup/benchmark/corrected/DecontX/",rep,"/resDefault_emptyTrue_cormat.RDS")),
                      SoupX = readRDS(paste0("Snakemake_backup/benchmark/corrected/SoupX/",rep,"/res1_contAuto_cormat.RDS")))
  
  marker_df <- lapply(cormat_list, function(cor_mat){
     cor_mat[topGenes,] %>% as.matrix %>% t() %>% data.frame() %>% rownames_to_column("cell") %>% 
      pivot_longer(cols = -cell, names_to = "gene", values_to = "expr") %>% 
      mutate(log_expr = log10(expr+1)) %>% 
      inner_join(cell_sample)
  }) %>% bind_rows(.id = "method")
  return(marker_df)
}) %>% bind_rows(.id = "replicate")
  

p.PT_heatmap_methods <- marker_plot_inp %>% 
  mutate(method = factor(method, levels = c("raw", "CellBender", "SoupX", "DecontX_default", "DecontX_background")),
         replicate = factor(replicate, levels = c("rep3", "rep1", "rep2", "nuc2")),
         gene = factor(gene, levels = rev(topGenes)),
         isPT = factor(isPT, levels = c("PT", "not PT"))) %>% 
  ggplot(aes(x = cell, y = gene, fill = log_expr))+
          geom_tile()+
          ggh4x::facet_nested(method~replicate+isPT, scale = "free", space = "free")+
          #facet_grid(~isPT, scales = "free", space = "free_x")+
          theme_bw()+
          scale_fill_gradient(low = "lightgrey", high = "red")+
          scale_x_discrete(position = "top")+ 
          theme(legend.position = "top", axis.text.y = element_text(size = 7), axis.title = element_blank(),
                legend.box.margin=margin(-10,-10,-10,-10), axis.text.x = element_blank(), axis.ticks.x = element_blank(),
                strip.background = element_rect(fill = "white"), strip.text.y =  element_text(size = 7))

```

### S13B: Overlap of top10 genes with Panglao markers

```{r panglao_suppl1}
panglao_markers <- readRDS("files/panglao_markers_Mm.RDS") 
sigSEU<- readRDS("Extra_analyses/DE_seurat_sigUP.RDS")  %>% 
  group_by(method2,replicate) %>% 
  mutate( rnk = dense_rank(-avg_log2FC),
          panglao = gene %in% panglao_markers) %>% 
  group_by(replicate,gene) %>% 
  mutate( Minrnk = min(rnk)) %>% ungroup 

sigSEU$method2<-factor(sigSEU$method2, levels=c("raw","CellBender","SoupX","DecontX_default","DecontX_background"))

top10.seu_panglao<-sigSEU %>% filter( panglao & rnk<=10) %>% 
  ggplot(aes(x=method2,  fill=method2)) +
  geom_bar() +
  scale_fill_manual(values=method_colors2, limits = force)+
  facet_grid(. ~ replicate) + coord_flip() +theme_bw()+ 
  ylab("overlap of top10 LFC genes with panglao markers")+
  theme(axis.title.y = element_blank(), legend.position =  "None")
```

### S13C: LFC of Panglao markers

```{r panglao_suppl2}
tmp<-sigSEU %>% filter(Minrnk<=100) %>% dplyr::transmute(method=method2, replicate, gene, panglao, avg_log2FC)
  
tmp2<-inner_join( tmp %>% filter(method !="raw") , 
            tmp %>% filter(method == "raw") %>% dplyr::select(-method),
            by = c("replicate","gene","panglao"),
            suffix = c("","_raw"))

top100cor<-tmp2 %>% ggplot( aes(x=avg_log2FC_raw, y=avg_log2FC,col=method)) +
  geom_abline(intercept=0,slope=1,lty=1,col="black")+
  geom_point(alpha=0.5)+ stat_smooth(method = "lm",se = F)+
  scale_color_manual(values=method_colors2, limits = force)+
  facet_grid(. ~ replicate) + theme_bw()+
  xlab("log2-fold-change uncorrected") +
  ylab("log2-fold-change corrected")+
  theme(legend.position = "bottom", legend.title = element_blank())

```

### Assemble Suppl Figure S13

```{r suppl_figure_markers, fig.width=10, fig.height=11}
p.topMarkers_suppl <- plot_grid(p.PT_heatmap_methods,
                                top10.seu_panglao,
                                top100cor,
          nrow = 3, rel_heights = c(2.3,0.8,1), labels = c("A","B","C"),
          align = "v", axis = "l")

p.topMarkers_suppl

ggsave("figures/figures_final/Suppl_figure_S13_topMarkersEval.pdf", p.topMarkers_suppl, width = 10, height = 11)
```

## SUPPL. FIGURE S14: Absolute evaluation metrics

```{r absolute_metrics, fig.width=9, fig.height=2.5}
p.absolute_clustering <- inp %>% 
  mutate(scaled_range = case_when(metric %in% c("avg_silhouette","prediction_score") ~ delta/2,
                                  metric %in% c("purity") ~ delta,
                                  metric %in% c("kNN_overlap") ~ delta/50,)) %>% 
  filter(evaluation_category == "cluster_evaluation") %>% 
  mutate(metric_long = factor(metric_long, levels = c("Prediction score (delta)", "Average silhouette","Purity","k-NN overlap")),
          replicate = factor(replicate, levels = rev(c("rep3", "rep1", "rep2", "nuc2")))) %>% 
  ggplot(aes(x = replicate, y = value, color = method2))+
      geom_jitter(size = 2, width = 0.1)+
      coord_flip()+
      scale_color_manual(values = method_colors2, limits = force, breaks = c("raw", "CellBender","SoupX", "DecontX_default", "DecontX_background"))+
      ggh4x::facet_nested(~metric_long, scale = "free")+
      theme_bw()+
      #labs(y = "scaled difference to score(uncorrected)")+
      theme(axis.title = element_blank(), axis.text.x = element_text(size = 6), strip.background = element_rect(fill = "white"),
                legend.title = element_blank())

p.absolute_clustering

ggsave("figures/figures_final/Suppl_figure_S14_metrics_absolute.pdf", p.absolute_clustering, width = 9, height = 2.5)
```

## SUPPL. FIGURE S15: Evaluation of different parameter settings

```{r param_settings, fig.width=8.5, fig.height=5.5}
df_zscore <- df_eval_metrics %>% 
  # Reverse metrics for which low value = better
  mutate(delta = ifelse(metric %in% c("rmsle", "expression_fraction"), -delta, delta)) %>% 
  # Remove estimation accuracy for parameter settings in which the contamination fraction has to be set  
  filter(!(evaluation_category == "estimation_accuracy" & grepl("cont0",param))) %>% 
  group_by(metric, replicate) %>% 
  mutate(mean=mean(delta, na.rm = T),
         sd=sd(delta, na.rm = T),
         dist_to_mean = delta-mean,
         zscore = dist_to_mean/sd) 

df_zscore_mergeReplicates <- df_zscore %>%
  group_by(method, param, evaluation_category, metric) %>% 
  summarise(mean_zscore = mean(zscore)) %>% 
  ungroup() %>% 
  group_by(metric) %>% 
  mutate(
    metric_available_for_raw = ifelse("raw" %in% method, T, F), 
    raw_value = ifelse(metric_available_for_raw == T, mean_zscore[method == "raw"], NA),
    improvement = ifelse(mean_zscore > raw_value, T, F),
    best_performance = ifelse(mean_zscore == max(mean_zscore), T, F))

inp_param_setting <- df_zscore_mergeReplicates %>% 
  filter(metric %in% c("tau", "rmsle","prediction_score", "purity", "avg_silhouette","expression_fraction", "lfc","kNN_overlap"),
         method != "raw", 
         !grepl("cont0.05",param)
         ) %>%
  mutate(param = sub("_total25000","",param),
         #param = sub("_emptyFalse", "", param),
         evaluation_category = case_when(evaluation_category == "estimation_accuracy" ~ "estimation \n accuracy",
                                         evaluation_category == "cluster_evaluation" ~ "cell type \n identification",
                                         evaluation_category == "marker_evaluation" ~ "marker gene \n detection"),
         evaluation_category = factor(evaluation_category, levels = c("estimation \n accuracy","marker gene \n detection","cell type \n identification")),
         param = ifelse(method == "SoupX", paste0(word(param,2,2,"_"),"_", word(param,1,1,"_")), param),
         param = factor(param,levels = c("fpr0.01","fpr0.05","fpr0.1",
                                         "resDefault_emptyFalse","res0.5_emptyFalse","res1_emptyFalse","res2_emptyFalse",
                                         "resDefault_emptyTrue","res0.5_emptyTrue","res1_emptyTrue","res2_emptyTrue",
                                         "contAuto_res1","contAuto_res0.5","contAuto_res2",
                                         "cont0.1_res0.5","cont0.1_res1","cont0.1_res2",
                                         "cont0.2_res0.5","cont0.2_res1","cont0.2_res2")),
         metric_pretty = case_when(metric == "expression_fraction" ~ "Unspecific \n detection",
                                   metric == "prediction_score" ~ "Prediction \n score",
                                   metric == "purity" ~ "Purity",
                                   metric == "avg_silhouette" ~ "Average silhouette",
                                   metric == "tau" ~ "Kendall's tau",
                                   metric == "rmsle" ~ "RMSLE",
                                   metric == "lfc" ~ "Log2 fold change",
                                   metric == "kNN_overlap" ~ "k-NN overlap",
                                   T ~ metric))

# highlight default in axis text
p.parameter_settings <-  ggplot(inp_param_setting, aes(x = param, y = metric_pretty, fill = mean_zscore))+
    geom_tile(color = "darkgrey")+
    #geom_boxplot(aes(fill = method), alpha = 0.7)+
    scale_fill_gradient2(low = "darkorange", high = "cyan4")+
    scale_y_discrete(position = "right")+
    facet_grid(evaluation_category~method,scales = "free", space = "free", switch = "y")+
    geom_hline(yintercept = 0)+
    theme_bw()+
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, color = c("red",rep("black",10))), 
          axis.title = element_blank(),strip.text.y = element_text(angle = 180), 
          legend.position = "top", legend.text = element_text(size = 8),
          strip.background = element_rect(fill = "white"), strip.text.x = element_text(face = "bold"))

p.parameter_settings

ggsave("figures/figures_final/Suppl_figure_S15_parameter_settings.pdf", width = 8.5, height = 5.5)
```